Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-07 analysis/paper/paper.Rmd
some final edits to Intro

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. Using a sensitivity analysis, Northern Pikeminnow predation on juvenile Chinook Salmon was predicted to reduce adult returns by 377 (95% CI: 161 - 935). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough contains historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Therefore, quantifying the mortality of spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG) associated with piscine predation is of particular interest. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore a prioritized candidate for restoration of natural processes [@Zimmerman2012].

The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as it supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

Recent studies have examined the downstream movement, distribution, and apparent survival (hereafter 'survival') of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had an approximate 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during fall and early winter months [@Ackerman2018; @Porter2019]. Low survival was attributed to lack of fish cover and low-velocity water delaying movement rates, thereby increasing predation risk.

@@

1. Estimate the abundance of Northern Pikeminnow in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3. Use an established bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4. Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough is within a migration corridor for several Chinook Salmon populations as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough, an approximately 1.9 kilometer reach of the Salmon River, to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. Using a sensitivity analysis, estimated Northern Pikeminnow predation on juvenile Chinook Salmon was equivalent to 377 (95% CI: 161 - 935) returning adults. Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impact on Chinook Salmon recovery in the Upper Salmon River.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Therefore, quantifying the mortality of spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon River associated with piscine predation is of particular interest. Importantly, the Salmon River supports eight extant populations of Chinook Salmon upstream of Deadwater Slough [@NOAA2017]. This includes the Lemhi River, which was historically the largest population, and is a prioritized candidate for restoration of natural processes to increase production of juvenile Chinook Salmon [@Zimmerman2012].

Chinook Salmon in the Upper Salmon River are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for Chinook Salmon in the Upper Salmon River as it supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

Recent studies have examined the downstream movement, distribution, and apparent survival (hereafter 'survival') of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have 32% lower survival in the Deadwater Slough reach compared to directly adjacent reaches [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had an approximate 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during fall and early winter months [@Ackerman2018; @Porter2019]. Low survival was attributed to lack of fish cover and low-velocity water delaying movement rates, thereby increasing predation risk.

@@

1. Estimate the abundance of Northern Pikeminnow in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3. Use an established bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4. Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon River.
